Is Vagueness Beneficial for Hearers?\\Evidence from Experiments

### 2016-09-17 papers/2016-Special-Issue/v-special-issue-2016.tex
holding point

It is, however, questionable whether the ``evaluative categories" manipulation in this study can be considered a manipulation of vagueness. Certainly, terms like \emph{Fair} admit the possibility of borderline cases. However, given that the boundaries between the categories were marked crisply, and that therefore the categories mapped crisply to numerical values, it becomes doubtful whether any borderline cases could be conceived to arise in fact. For example, \emph{Fair} was mapped to 60\% -- 70\% for the variable \emph{percentage of heart attack patients given recommended treatment (ACE inhibitor)}. Accordingly, rather than the vagueness of categories such as \emph{Poor}, Peters et al. emphasise the evaluative content inherent in these categories, and the affective potential of the evaluative content rather than the vagueness of the terms like {\em Fair}.
%\\[2ex]
\subsection{}
The experiments reported in the present paper put the cost reduction hypothesis to the test. The question that we are trying to answer is whether vague expressions are processed more easily by readers than crisp ones. Like Lipman, we focus on situations where numerical information is used in order to identify a referent. Reference, in other words, will be the linguistic task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

@@

In our experiments we used a speeded forced choice task to compare the processing costs of different references to quantities. In this context, speed and accuracy of responses are the key dimensions on which the different references can be compared. The stimuli in the experiments were sets of dot arrays containing various numbers of dots. The forced choice was to identify one dot array given a reference to a given quantity of dots. We manipulated the references in several ways across a series of four experiments.

Our main manipulation was always of vagueness: we constructed crisp and vague versions of references to the same dot array. For example, in experiment 1 the instruction presented to the participant could be \emph{Choose the square with many dots.} (vague condition), or \emph{Choose the square with 20 dots.} (crisp condition), identifying the same dot array.

Ideally, we would manipulate vagueness independently of other variables. However, in practice, and given the constraints imposed by the use of natural language, any manipulation of vagueness introduces variance along other dimensions as well. Therefore we carried out further experiments to try to address and control variance along other dimensions.

One non-vagueness source of variance was the difference between numerical and verbal format in the references. For example, in the instructions for experiment 1 given above, \emph{many} is in verbal format (as well as vague) and \emph{20} is in numerical format (as well as crisp). Therefore any difference observed between responses to the \emph{many} instruction and the \emph{20} instruction could be due either to vagueness or to {\bf instruction format}. It is possible to create vague and crisp references in each of these instruction formats: for example in numerical instruction format we can have \emph{20} (crisp) and \emph{about 20} (vague), and in verbal instruction format we can have \emph{the most} (crisp) and \emph{many} (vague). Experiment 2 varied vagueness and instruction format factorially in a similar way to this, to try to tease apart these two sources of variance.

Another potential non-vagueness source of variance between responses to the \emph{many} (verbal vague) references and the \emph{20} (numeric crisp) references lies in the method used to identify a referent (we call this the {\bf selection algorithm} source of variance). Identifying the array with \emph{many} dots might be done by a \emph{comparison} algorithm, identifying that one array is more numerous than the others without establishing the numerosity of any arrays, whereas identifying the array with \emph{20} dots, which might be done by a \emph{matching} algorithm would require an estimate of the numerosity of the arrays. It seems reasonable that comparison would be faster than matching because it does not require estimates of cardinality. Furthermore the problem persists when using references like \emph{the most} (verbal crisp) and \emph{about 20} (numeric vague), since they too vary not only along the vague/crisp dimension and the numeric/verbal dimension but also along the comparison/matching dimension. Experiments 3 and 4 addressed the \emph{selection algorithm} source of variance - experiment 3 crossed vague/crisp and comparison/matching factorially using only numerical references, and experiment 4 crossed vague/crisp and comparison/matching factorially using only verbal references.

@@

\subsection{Introduction}
We used a forced choice task to compare choices made in response to vague instructions against choices made in response to crisp instructions. The participant was presented with an instruction like ``Choose the square with many dots'' in the vague conditions, or ``Choose the square with 20 dots'' in the crisp conditions. Then two dot arrays were presented in the form of squares containing a number of dots. Fig \ref{stimuluse1} shows an example stimulus. The participant was required to identify the square that corresponded with the instruction, by pressing the appropriate key.
Response time and accuracy were recorded for analysis.

Our main manipulation was of the vagueness of the instruction, with two levels, vague and crisp. Table \ref{instructionse1} shows examples from each condition.  We also manipulated how discriminable the dot arrays were. One array always contained 25 dots: the other contained either 5, 10, 15, 20, 30, 35, 40, or 45 dots. This led to numerical differences of 5, 10, 15, and 20, with lower differences resulting in less discriminable arrays and larger differences representing more discriminable arrays. There is evidence that when the distance grows between two numbers, they become more easily distinguishable from each other: the \emph{numerical distance effect}, which has been shown for comparing the numerosity of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene199647}. Where a number was mentioned in the instruction, it was always in the form of an Arabic numeral.

The instructions indicated the larger of the two dot arrays equally often as they indicated the smaller of the dot arrays, so that participants could not systematically choose the larger or smaller array as a successful response strategy. There is evidence that when two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

\subsection{Hypotheses}
The {\bf cost reduction} hypothesis predicts that responses will be faster and more accurate for the vague conditions than for the crisp conditions because the vague conditions impose a lower cognitive load than the crisp conditions.

The {\bf instruction format} hypothesis predicts that responses will be faster and more accurate in the vague (i.e., verbal) conditions than in the crisp (i.e., numerical) conditions because numbers are harder to process than verbal references to quantities.

The {\bf selection algorithm} account predicts that responses will be faster and more accurate in the vague (i.e., comparison) conditions than in the crisp (i.e., matching) conditions because it is easier to carry out comparison than matching.

Thus all three accounts predict a main effect advantage of vagueness, but for different reasons. All three accounts also predict a main effect advantage for greater numerical distance, because greater numerical distances lead to more easily discriminable arrays. The accounts also share the prediction of an interaction effect whereby the main effect advantage of vagueness should diminish as numerical distance increases.

% KvD I'm tentatively starting with a general explanation of the methodology of this group of experiments.

%{\bf General.} All our experiments shared essentially the same methodology: Participants aged between ..., self-reported fluent speakers of English, were recruited by .... and were paid 10 ponds for participating. All had normal, or corrected-to-normal vision. A MacBook ... Stimuli were created ... In all except the last experiment (Experiment 5), items were presented as follows: (...) (IF WE DO ADD THIS GENERAL DESCRIPTION THEN REMOVE THE RELEVANT INFO FROM THE DESCRIPTION OF EACH OF THE INDIVIDUAL EXPERIMENTS.)

\subsection{Method}
Twenty participants were recruited by mailing list and paid ten pounds for participating. Participants were aged between 18 and 45, with a median age of 26. All participants self-reported fluency in English, and had normal, or corrected-to-normal vision. A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants. Stimuli were created and presented using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}. We needed to control which side the target was presented on. Each pair was presented equally often with with the target on the left, as with the target on the right. We also needed to control whether the target was the smaller or larger number. The items with numbers less than 25 (pairs 1 to 4) formed a group with the smaller number as target and the other items (pairs 5 to 8) formed a balancing group with the larger number as the target number.

The experiment was conducted in a quiet room. On arrival in the room, the participant was told that he or she would be presented with an instruction to choose one of two squares by reference to how many dots it contained. Participants were required to press the space key after reading the instruction. Then there was a central fixation cross for 1000 ms, and a blank screen for 500 ms, followed by the squares and dots (without repetition of the referring expression).  The position of the dots was randomised per-trial. Response time was measured as the latency between the presentation of the dots and squares, and the keypress identifying the decision; in this way, the response times can be separated from time spent reading the instructions, which is important since we are only interested in the former.

The display would stay on screen until the participant responded (there was no timing-out). Participants were asked to respond quickly while avoiding errors. There were 8 practice trials, after which the participant was invited to ask any questions about procedure. After answering these the experimenter left the cubicle for the duration of the experiment. The order in which trials were presented was randomised per-participant. There were 256 trials, presented in 4 blocks of 64 trials each, between which the participant could rest. No feedback was given on correct trials, but there was feedback on error trials in the form of the word ``\textsc{wrong!!}'' which flashed on screen.

\subsection{Results}

A response was counted as erroneous if the square with the wrong number of dots was chosen (when the instruction contained a number); if the square with the larger number of dots was selected (when the instruction was ``Choose the square with few dots"); or if the square with the smaller number of dots was selected (when the instruction was ``Choose the square with many dots".) % KvD Is this addition defining "erroneous" OK?
RTs for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of 160 trials, 3.4\% of the correct responses. Means for response times and error rates are given in Fig. (\ref{resultse2}).
A linear mixed model of RT was built using as independent variables vagueness and gap size and their interaction, with random slopes for vagueness and gap size over participants. Vagueness was sum coded: vague $= -.5$, crisp $= .5$; gap size was Helmert coded. Helmert contrasts compare each level against the mean of the previous levels. Level one of this contrast is gap size 5 compared with gap size 10; level two is the mean of gap size 5 and gap size 10 versus gap size 15; and level three is the mean of gap sizes 5, 10 and 15 versus gap size 20. 
$p$ values were calculated using the R package \emph{lmerTest} \cite{lmerTest}.

RTs were faster for vague instructions ($\beta=.109, se=.022, t=4.9, p<.001$). 
%
RT grew faster as gap size increased: level one $(\beta=-.116, se=.012, t=-9.3, p<.05)$, level 2 ($\beta=-.103, se=.009, t=-11.1, p<.001$) and level three ($\beta=-.082, se=.007, t=-11.0, p<.05$). Since discriminability of the dot arrays is easier for larger gap sizes, discriminability probably underlies this effect. Gap size and vagueness interacted significantly for larger gap sizes when modelling RT. % KvD Is it worth saying how they interacted?
The interactions at the different levels of gap size were: level one: ($\beta=.001, se=.016, t=.03, p=.974$); level two ($\beta=-.039, se=.009, t=-4.460, p<.001$); level three ($\beta=-.043, se=.006,t=-7.0, p<.001$). In the crisp conditions RTs started out much slower than in the vague conditions, at the smallest gap size, but the two conditions converged to very fast times at the largest gap size. There were diminishing returns for vagueness as gap size increased.

\label{accann}
Error rate data were analysed using a generalized logit mixed model \cite{jaeger2008categorical}, with vagueness and gap size and their interaction as independent variables, and with random slopes for vagueness and gap size over participants. 
%
The effect of vagueness on error rates approached significance, with the vague instructions leading to fewer errors ($\beta=.307, se=.173, t=1.8, p=.077$).
%
Error rates decreased as gap size increased: level one ($\beta=-.585, se=.092, ,z=-6.4, p<.001$), level 2 ($\beta=-.434, z=-5.3,p<.001$) and level three ($\beta=-.250,z=-4.1, p<.001$). 
%
Error rates were greater in the crisp conditions than the vague conditions when gap size was small, and this difference diminished with increasing gap size until it reversed at the biggest gap size.
\begin{figure}[htbp]
\fitfigure{images/resultse1.pdf}
\caption{Experiment 1 results. }
\label{resultse1}
\end{figure}

@@

Results from experiment 1 were in line with the three main accounts. Responses were faster and more accurate when the instructions were in the vague (or verbal, or comparison) conditions than when they were in the crisp (or numerical, or selection) conditions. The speed and accuracy diminished as the arrays became more easily discriminable, and the advantage for vague (or verbal, or comparison) instructions tailed off as the arrays became more easily discriminable.

The cost reduction hypothesis explains the vagueness advantage by claiming that the vague referring expressions place less cognitive load on the comprehender than the crisp referring expressions. It explains the diminishing returns for vagueness in more-discriminable stimuli by claiming that load is low in both conditions for the easily-discriminable stimuli, and that therefore there is no extra benefit to be had from vagueness in the easily-discriminable stimuli.

The instruction format hypothesis explains the main effect advantage for the vague instructions by observing that the vague instructions used verbal quantifiers whereas the crisp conditions used numerical quantifiers. Under this account it is avoiding numbers that explains the vagueness advantage main effect. The main effect of numerical distance is explained by assuming that larger distances result in more easily discriminable arrays. The vagueness x numerical distance interaction can be explained by assuming that the numerical quantifiers make the task particularly challenging when the stimuli are less discriminable.

The selection algorithm account explains the main effect advantage of the vague conditions as due to the vague conditions allowing a comparison strategy rather than a matching strategy in the crisp conditions. The main effect of numerical distance is explained by claiming that comparison is easier for more discriminable arrays. The diminishing returns for vagueness as numerical distance grows are explained by claiming that the advantage of being able to use comparison is greater for less discriminable arrays than for more discriminable arrays.

\section{Experiment 2}

The main result from experiment 1 was that responses were faster and more accurate for vague instructions than for crisp instructions. This finding can be interpreted in line with three different accounts: the cost reduction account, the instruction format account, and the selection algorithm account.

In experiment 2 we set out to distinguish between the cost reduction account and the instruction format account, leaving the selection algorithm account for experiments 3 and 4. Contrast an expression from the vague condition: `the square with few dots' with an expression from the crisp condition: `the square with 15 dots'. One difference is that `few' is vague (or at least has the potential for vagueness) and `15' is crisp. Another difference is that `few' is a linguistic, or verbal quantifier while `15' is a numerical quantifier, in the sense that a number is mentioned explicitly. Since these two differences were confounded in Experiment 1, the vagueness advantage finding is vulnerable to an alternative interpretation, that what we saw as a vagueness advantage was in contrast an advantage for the verbal form of the quantifier. In the present experiment 2 we pitted these alternative interpretations against each other in a factorial design.

In Experiment 1, the participants chose one of two squares. The `vague' quantifiers (e.g., `few') uniquely identified one square. Recall our definition of vague  -- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''.  In the first experiment, the quantifiers in the vague conditions did not really meet this definition. This is because there were no borderline cases of the referent that could make the referent set `not well-defined'. Experiment 2 used three squares so that the vague quantifiers always had more than one possible referent. To enhance the potential for the vague quantifiers to have true vagueness in Experiment 2, we also used indefinite articles in the vague instructions. Thus, instead of the instruction \emph{Choose the square with few dots}, we used \emph{Choose a square with few dots}.

In experiment 2, an item was a referring expression followed by a triple of numbers, representing the number of squares in the left, middle, and right squares. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each triple had the following properties: it comprised three squares (instead of two as in Experiment 1); the central number was always presented in the middle of the three; there were two flanking numbers where one was smaller than the central number and one was bigger.

There was a numerical and a verbal version of each of the vague and crisp referring expressions. See Table \ref{instructionse2} for examples.
The {\em vague numerical condition}'s referring expression was, on half of the presentations in that condition, ``Choose a square with about 10 dots''. None of the squares displayed contained 10 dots. 10 is slightly closer to 6 than to 15. Therefore the best referent for this referring expression was the square with 6 dots;  the borderline response was the square with 15 dots; and the poorest referent was the square with 24 dots. On the other half of the presentations in that condition the referring expression was ``Choose a square with about 20 dots''. The {\em crisp numerical} conditions's referring expression was ``Choose the square with 6 dots'' on half the presentations in that condition and ``Choose the square with 24 dots'' on the other half. One square did contain the exact number mentioned.  In the {\em crisp verbal} condition, we used the referring expression ``Choose a square with fewer than $20$ (more than $10$) dots".  The {\em vague numerical} condition's referring expression was ``Choose the square with far fewer than $20$ dots'' on half of presentations, making the squares with 6 and $15$ dots possible referents; in the other case, the referring expression was ``Choose the square with far more than $10$ dots'', which made $15$ and $24$ possible referents.The {\em vague verbal} condition's referring expression was ``Choose a square with few dots'' on half of the presentations, and ``Choose a square with many dots'' on the other half. For this condition, the best referent was the square with 6 dots for `few' and 24 dots for `many'; the borderline case was the square with 15 dots; and the remaining square was the poorest referent.

An indication that the manipulation of vagueness was successful is that participants chose the borderline case square on 16\% of trials.

\begin{table}[htbp]
\caption{Table of instructions arranged by condition for the triple (6,15,24). Experiment 2}
\label{instructionse2}
\begin{tabular}{rlll}
\toprule
vagueness&instruction format&selection task&example\\
\midrule
crisp 	& 	numeric	& matching	&Choose the square with 6 dots \\
vague	&	numeric & matching	&Choose a square with about 10 dots\\
crisp	&	verbal	& comparison	&Choose the square with the fewest dots\\
vague	&	verbal	& comparison	&Choose a square with few dots\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Hypotheses}
The cost reduction account predicts a main effect of vagueness such that the vague instruction conditions attract faster responses than the crisp instruction conditions. The cost reduction account also predicts an interaction with the instruction format (or selection algorithm) variable: at each level of instruction format (or selection algorithm) the vague condition should attract faster responses than the crisp condition.%why is this interaction predicted?

>>>>

It is, however, questionable whether the ``evaluative categories" manipulation in this study can be considered a manipulation of vagueness. Certainly, terms like \emph{Fair} admit the possibility of borderline cases. However, given that the boundaries between the categories were marked crisply, and that therefore the categories mapped crisply to numerical values, it becomes doubtful whether any borderline cases could be conceived to arise in fact. For example, \emph{Fair} was mapped to 60\% -- 70\% for the variable \emph{percentage of heart attack patients given recommended treatment (ACE inhibitor)}. Accordingly, rather than the vagueness of categories such as \emph{Poor}, Peters et al. emphasise the evaluative content inherent in these categories, and the affective potential of the evaluative content rather than the vagueness of the terms like {\em Fair}.
\clearpage
%\\[2ex]
\subsection{General Methodology}
The experiments reported in the present paper put the cost reduction hypothesis to the test. 
The question that we are trying to answer is whether vague expressions are processed more easily by readers than crisp ones. 
Like Lipman, we focus on situations where numerical information is used in order to identify a referent. 
Reference, in other words, will be the linguistic task on which we focus, partly because of the interest that this topic has recently drawn from the NLG community.  
In focussing on benefits for the hearer, we will leave aside the question of audience design, leaving this for later research.

@@

In our experiments we used a speeded forced choice task to compare the processing costs of different references to quantities. 
In this context, speed and accuracy of responses are the key dimensions on which the different references can be compared.

Each stimulus in the experiments was a set of dot arrays containing various number of dots, together with a preceding instruction (in the form of a referring expression) to choose one of the arrays with respect to its number of dots, as quickly as possible while avoiding errors. 
We manipulated the stimuli in several ways across a series of four experiments.

% added/moved out of expt 1 by matt 17 sept
There is evidence that when the distance grows between two numbers, they become more easily distinguishable from each other: the \emph{numerical distance effect}, which has been shown for comparing the numerosity of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene199647}. 
Where a number was mentioned in the instruction, it was always in the form of an Arabic numeral.

% added/moved out of expt 1 by matt 17 sept
There is evidence that when two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. 
We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

% General methodology moved to here
All the experiments shared the following properties: 
The experiment was conducted in a quiet room.
All participants self-reported fluency in English, and had normal, or corrected-to-normal vision.
Participants were asked to respond quickly while avoiding errors.
Stimuli were created and presented using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}.
There was a block of practice trials after which participants could ask any questions, following which the experimenter left the room.
The position of the dots was randomised per-trial.
The order in which trials were presented was randomised per-participant.
A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants.

\clearpage

@@

\subsection{Introduction}
We used a forced choice task to compare choices made in response to vague instructions against choices made in response to crisp instructions. 
The participant was presented with an instruction like ``Choose the square with many dots'' in the vague conditions, or ``Choose the square with 20 dots'' in the crisp conditions. 
Then two dot arrays were presented in the form of squares containing a number of dots. %Fig \ref{stimuluse1} shows an example stimulus. 
The participant was required to identify the array that corresponded with the instruction, by pressing the appropriate key.
Response time and accuracy were recorded for analysis.

Our main manipulation was of the vagueness of the instruction, with two levels, vague and crisp. %Table \ref{instructionse1} shows examples from each condition.  
We also manipulated how discriminable the dot arrays were. 
One array always contained 25 dots: the other contained either 5, 10, 15, 20, 30, 35, 40, or 45 dots. 
This led to numerical differences of 5, 10, 15, and 20, with lower differences resulting in less discriminable arrays and larger differences representing more discriminable arrays.

\subsection{Hypotheses}
(1) A main effect of vagueness: vague instructions impose a lower cognitive load for the comprehender than crisp alternatives.

(2) A main effect advantage of increasing numerical distance: the task will become easier as the numerical distance increases, because the two arrays are more discriminable.

(3) An interaction between vagueness and numerical distance: i.e., any facilitation for vagueness should be greater at smaller numerical distances.

\subsection{Method}
Twenty participants were recruited by mailing list; aged between 18 and 45 with a median age of 26; and paid ten pounds for participating. 
On arrival in the room, each participant was told that on each trial he or she would be presented with an instruction to choose one of two squares on screen by reference to how many dots the square contained. 
Participants were required to press the space key after reading the instruction. 
Then there was a central fixation cross for 1000 ms, and a blank screen for 500 ms, followed by the array (without repetition of the referring expression). 
The arrays would stay on screen until the participant responded (there was no timing-out). 
Response time was measured as the latency between the presentation of the arrays, and the keypress identifying the choice: in this way, the response time was separated from time spent reading the instructions, which is important since we are only interested in the former. 
There were 256 trials, presented in 4 blocks of 64 trials each, between which the participant could rest. 
No feedback was given on correct trials, but there was feedback on error trials in the form of the word ``\textsc{wrong!!}'' which flashed on screen.

\subsection{Results}
A response was counted as erroneous if the square with the wrong number of dots was chosen (when the instruction contained a number); if the square with the larger number of dots was selected (when the instruction was ``Choose the square with few dots"); or if the square with the smaller number of dots was selected (when the instruction was ``Choose the square with many dots".) % KvD Is this addition defining "erroneous" OK?
RTs for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. 
The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of a further 160 trials, or 3.4\% of the remaining correct responses. 
Means for response times and error rates are given in Fig. (\ref{resultse2}).
A linear mixed model of RT was built using as independent variables vagueness and numerical distance and their interaction, with random slopes for vagueness and numerical distance over participants. 
Vagueness was sum coded: vague $= -.5$, crisp $= .5$; numerical distance was Helmert coded. 
Helmert contrasts compare each level against the mean of the previous levels. 
Level one of this contrast is numerical distance 5 compared with numerical distance 10; level two is the mean of numerical distance 5 and numerical distance 10 versus numerical distance 15; and level three is the mean of numerical distances 5, 10 and 15 versus numerical distance 20. 
All $p$ values were calculated using the R package \emph{lmerTest} \cite{lmerTest}.

(1) RTs were faster for vague instructions than for crisp instructions ($\beta=.109, se=.022, t=4.9, p<.001$). 
(2) RTs grew faster as numerical distance increased: level one $(\beta=-.116, se=.012, t=-9.3, p<.05)$, level 2 ($\beta=-.103, se=.009, t=-11.1, p<.001$) and level three ($\beta=-.082, se=.007, t=-11.0, p<.05$). 
Since discriminability of the arrays is easier for larger numerical distances, discriminability probably underlies this effect. 
(3) Numerical distance and vagueness interacted significantly for larger numerical distances when modelling RT: essentially there were diminishing returns for vagueness as numerical distance increased.% KvD Is it worth saying how they interacted? % MG I think this is answered now?
The interactions at the different levels of numerical distance were: level one: ($\beta=.001, se=.016, t=.03, p=.974$); level two ($\beta=-.039, se=.009, t=-4.460, p<.001$); level three ($\beta=-.043, se=.006,t=-7.0, p<.001$). 
In the crisp conditions RTs started out much slower than in the vague conditions, at the smallest numerical distance, but the two conditions converged to very fast times at the largest numerical distance.

Error rate data were analysed using a generalized logit mixed model \cite{jaeger2008categorical}, with vagueness and numerical distance and their interaction as independent variables, and with random slopes for vagueness and numerical distance over participants. 
The effect of vagueness on error rates approached significance, with the vague instructions leading to fewer errors ($\beta=.307, se=.173, t=1.8, p=.077$).
Error rates decreased as numerical distance increased: level one ($\beta=-.585, se=.092, ,z=-6.4, p<.001$), level 2 ($\beta=-.434, z=-5.3,p<.001$) and level three ($\beta=-.250,z=-4.1, p<.001$). 
Error rates were greater in the crisp conditions than the vague conditions when numerical distance was small, and this difference diminished with increasing numerical distance until it reversed at the biggest numerical distance.

\begin{figure}[htbp]
\fitfigure{images/resultse1.pdf}
\caption{Experiment 1 results: response times and error rates}
\label{resultse1}
\end{figure}

\subsection{Discussion}
Responses were faster and more accurate when the instructions were in the vague conditions than when they were in the crisp conditions. 
The speed and accuracy diminished as the arrays became more easily discriminable, and the advantage for vague instructions tailed off as the arrays became more easily discriminable.

The cost reduction hypothesis explains the vagueness advantage by claiming that the vague referring expressions place less cognitive load on the comprehender than the crisp referring expressions. 
It explains the diminishing returns for vagueness in more-discriminable stimuli (i.e., the vagueness by numerical distance interaction) by claiming that load is low in both conditions for the easily-discriminable stimuli, and that therefore there is no extra benefit to be had from vagueness in the easily-discriminable stimuli.

\clearpage
\section{Experiment 2}

The main result from experiment 1 was that responses were faster and more accurate for vague instructions than for crisp instructions.

Contrast, from Experiment 1, an expression from the vague condition: `the square with few dots' with an expression from the crisp condition: `the square with 5 dots'. 
One difference is that `few' is vague (or at least has the potential for vagueness) and `5' is crisp. 
Another difference is that `few' is a linguistic, or verbal quantifier while `5' is a numerical quantifier, in the sense that a number is mentioned explicitly. 
Since these two differences could not be separated in Experiment 1, the vagueness advantage finding is vulnerable to an alternative interpretation, that what we saw as a vagueness advantage was in contrast an advantage for the verbal form of the quantifier. 
In Experiment 2 we created verbal and numeric versions of each of the vague and crisp instructions so that we could compare vague and crisp conditions while taking account of verbal/numeric format.%In the present experiment 2 we pitted these alternative interpretations against each other in a factorial design.

In Experiment 1, the participants chose one of two squares: therefore the `vague' quantifiers (e.g., `few') uniquely identified one square. 
Recall our definition of vague  -- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''.  
In Experiment 1, it might be said that the quantifiers in the vague conditions did not realise their potential for vagueness. 
This is because there were no borderline cases of the referent that could make the referent set `not well-defined', and perhaps because using definite articles in the instructions implied that only one option was correct.
Using error feedback in experiment 1 could have exacerbated this.
Experiment 2 used three squares so that the vague quantifiers always had more than one possible referent, and used indefinite articles in the vague instructions to avoid the impression that only one response counted as correct, and was carried out without error feedback.
An indication that the potential for vagueness was realised in Experiment 2 is that the borderline response was chosen fairly often --16\% of the time.

In Experiment 2, an item was a referring expression followed by a triple of numbers, representing the number of squares in the left, middle, and right squares. 
We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). 
Each triple had the following properties: it comprised three squares (instead of two as in Experiment 1); the central number was always presented in the middle of the three; there were two flanking numbers where one was smaller than the central number and one was bigger.
Examples of crisp and vague versions of the numerical and verbal instructions follow: the examples assume the triple (6,15,24) and reference to the smaller numbers of dots, such that 6 was classified as the expected response; 15 was classified as the borderline response; and 24 was classified as the incorrect response. 
In the {\em vague numerical} condition we used ``Choose a square with about 10 dots''. None of the squares displayed contained 10 dots. 10 is slightly closer to 6 than to 15, justifying 6 as the best response and 15 as the borderline response.
In the {\em vague verbal} condition we used ``Choose a square with few dots''.
In the {\em crisp numerical} condition we used ``Choose the square with 6 dots''. One square did contain the exact number mentioned.
In the {\em crisp verbal} condition, we used ``Choose the square with the fewest dots".

\subsection{Hypotheses}
(1) A main effect advantage of vagueness.

(2) An advantage of vagueness both in \emph{numeric} and in \emph{verbal} instructions.

(3) No large main effect of instruction format. %MG If we are right that vagueness was responsible for the vagueness advantage in e 1, that could also have been a number avoidance advantage, then we would not expect that number avoidance, or, instruction format, plays a role.

(4) On the basis of Experiment 1, we would expect an effect of array, with faster responses to more discriminable arrays

Participants were aged between 18 and 45 with a median age of 28. 
We presented participants with 256 trials, arranged in 4 blocks each with 64 trials. 
We manipulated as independent variables vagueness and instruction format, yielding four conditions, \emph{vague numeric}; \emph{vague verbal}; \emph{crisp numeric}; \emph{crisp verbal}.
We measured two dependent variables: response time; and the probability of a participant choosing the borderline case.
On each trial, first the referring expression that constituted the instruction for that trial was displayed. 
The participant then pressed a key to indicate that he or she had read the instruction. 
After 1 second, the squares and dots were presented, while preserving the text of the referring expression. 
The response time dependent variable was measured from the presentation of the squares and dots, until the keypress indicating the participant's choice. 
The dependent variable measuring whether the participant chose the borderline case was also recorded at this time. 
The trial would timeout after 60 seconds if there was no response.
In this experiment, no feedback was given. 
This was because, in the vague conditions, we did not regard any response as `correct' or `incorrect', but instead as `borderline response', or `not borderline response', and we did not want to draw participants' attention to this distinction explicitly. 
We simply recorded whether the participant chose the best referent, the borderline case or the poorest referent, and how long it took the participant to respond.

\subsection{Results}
Means for response times and proportion of borderline responses are given in Fig. (\ref{resultse2}). 
Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. 
A linear mixed model was constructed for the (logged) response times.
The \emph{instruction format} and \emph{vagueness} variables were sum-coded and \emph{Item} was centred.
The fixed effects in the model were \emph{instruction format} and \emph{vagueness} and their interaction, and \emph{item}.
The random effects in the model were \emph{participant}, and slopes over \emph{participant} for \emph{instruction format}, \emph{vagueness}, and their interaction

(1) The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06, se=.01, t=4.6, p<.001$). 
(2) This disadvantage for vagueness persisted at each level of instruction format, offering evidence against hypothesis 2. [GIVE VALUES OF A FOCUSSED T TEST] There was an interaction effect between vagueness and instruction format, such that the disadvantage for vagueness was greater in the numerical than in the verbal conditions ($\beta=-0.13, se= 0.02, t= -6.6, p=0.000$).
(3) There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1 (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; $\beta=.37, se=.07, t=5.1, p<.001$). 
(4) There was a significant main effect of \emph{item} ($\beta=.06, se=.008, t=7.0, p<.001$) indicating that response times differed in some way according to which item was presented. However further analysis revealed that there was no consistent smooth trend across items. This effect seems likely to be due to the very fast responses for the smallest item. This item had the largest ratio difference so may have been particularly discriminable for participants. % MG reviews asked for an explanation of the difference for numeric crisp over items. Also reviews mention ratio differences as something that we did not control.

Participant grand mean percentage of borderline selections was $16.6\%$. A generalized linear mixed model \cite{jaeger2008categorical} was fit to the data for selection of the borderline response, with task, vagueness and item as fixed effects, and with random slopes for task and vagueness and item over participants.  
The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultse2}.
Participants were significantly more likely to choose the borderline option for vague instructions than for precise instructions (21.9\% vs 11.3\%, $\beta=.79, se=.25, z=3.2, p<.01$). 
Participants were significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%, $\beta=3.57, se=.26, z=13.6, p<.001$).

\begin{figure}[htbp]
\centering
\includegraphics[width=.49\textwidth]{images/response-time-sep-1.pdf}
\includegraphics[width=.49\textwidth]{images/histbd4-ggplot-version-1.pdf}
\label{resultse2}
\caption{Experiment 2 results: response times and proportion of borderline cases}
\end{figure}

@@

Experiment 2 tested to see whether when borderline cases are present, vague instructions would speed responses as they did in Experiment 1 when there were no borderline squares. 
We actually found a \emph{disadvantage} of vague instructions: vague instructions slowed people down by 112 ms on average. 
We also found that the effect of instruction format was significant, with numerical format slowing responses by 689 ms on average. 
The disadvantage of numerical format overwhelms the contribution of vagueness. 
The \emph{verbal vague} condition was still responded to faster than the \emph{numerical crisp} condition, so the pattern from Experiment 1 is reproduced, but in the light of the evidence from Experiment 2 in the presence of borderline cases, the advantage that was ascribed to vagueness before now looks more like either an advantage of verbal instruction format.

% KvD: I've tentatively added the following:
Having effectively separated the {\bf cost reduction} hypothesis from the {\bf instruction format} hypothesis, it is important to observe that, in Experiment 2, instruction format went hand in hand with {\bf selection algorithm}: 
 as shown by the Table of instructions \ref{instructionse2}, the instructions that used a verbal instruction format allowed a comparison algorithm, whereas the instructions that used a numeric format allowed a matching algorithm. 
Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison algorithm 
(which is known to be faster than matching). % KvD I assume we need to say more about this, citing some appropriate papers to support this idea.

In the next two experiments we pitted the comparison algorithm and matching algorithm selection tasks against each other while controlling vagueness and instruction format. 
In Experiment 3 we would restrict all the instructions to numeric quantifiers while factorially manipulating vagueness and selection task. 
In Experiment 4 we would ensure that all instructions used verbal quantifiers, while also factorially manipulating vagueness and selection task. 
This allows us to distinguish between the predictions of the selection task account and the instruction format account.

\clearpage
\section{Experiment 3}
